Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-10-24 prop_letter.tex
prop_letter.tex \section{Born rule and Gleason's Theorem} Updating...

As we have discussed, Statement~\ref{enu:iota-subject-to} implies
\ref{enu:iota-mu-QIVPM}. The rest of the proof is to show that if
$\iota$ is not a THOS rule, then $\iota\circ\mu$ might not be a
QIVPM. Suppose there are three particular numbers~$x_{0}$, $x_{1}$,
and $x_{2}\in\left[0,1\right]$ such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$,
but they doesn't satisfy Eq.~(\ref{eq:iota-Inclusion}). Consider
the state 
\begin{equation}
\begin{aligned}\rho={} & x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}\\
 & +\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}\,.
\end{aligned}
\end{equation}
whose inducing function $\iota\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}$
fails to satisfy Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}=\proj{0}+\proj{2}$ and $P_{1}=\proj{1}+\proj{2}$, and
cannot be a QIVPM. Therefore, $\iota$ is a THOS rule if and only
if $\iota\circ\mu$ is always a QIVPM. Similarly, Statement~\ref{enu:iota-subject-to}
and \ref{enu:iota-mu-CIVPM} are equivalent as well~\citep{HOSTunpublished}.

In contrast with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM},
there is always a state consistent with composited QIVPM $\iota\circ\mu_{\rho}^{\mathrm{B}}$.
Moreover, the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower. Eventually, the collection of consisting states collapses
to the singleton set~$\left\{ \rho\right\} $ as the intervals becomes
the infinite precise intervals~$\mathscr{I}_{\infty}$. This fact
is a weak version of Gleason's theorem because we only consider composited
QIVPMs $\iota\circ\mu_{\rho}^{\mathrm{B}}$ instead of general QIVPMs~$\bar{\mu}$.
To state this result, we need to introduce the norm of pure states
and operstors. The norm of a pure state~$\ket{\psi}$ is defined
as usual $\left\Vert \psi\right\Vert =\sqrt{\ip{\psi}{\psi}}$, and
the norm of the difference between two states~$\rho$ and $\rho'$
is just the usual operator norm
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert \left(\rho-\rho'\right)\ket{\psi}\right\Vert \,,
\end{equation}
which is also known as $2$-norm or spectral norm~\citep{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.
Then, a weak version of Gleason's theorem can be stated as follow.

@@

Given a state~$\rho'$ consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$,
we have $\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota\left(\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$
for any 1-dimensional projector~$\proj{\psi}$. Since the maximum
length of the intervals in $\mathscr{I}$ is $\delta$, it is also
the upper bound of the difference 
\[
\left|\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right|=\left|\melem{\psi}{\rho-\rho'}{\psi}\right|\le\delta\,.
\]
Since $\rho-\rho'$ is Hermitian, $\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$
is the maximum absolute value of the eigenvalues of $\rho-\rho'$~\citep{544199},
and equal to $\left\Vert \rho-\rho'\right\Vert $~\citep{GolubVanLoan1996,Foucart2012}.
Hence, $\left\Vert \rho-\rho'\right\Vert \le\delta$.

@@

Even if $\iota$ has to be a THOS rule to guarantee $\iota\circ\mu$
a QIVPM, the composited function $\iota\circ\mu$ may still be a QIVPM
when $\iota$ is not a THOS rule. In this case, the composited function
$\iota\circ\mu$ could map to only finite number of intervals. For
example, we can weaken the convex condition, Eq.~(\ref{eq:iota-Inclusion}),
to the following additive condition: for any numbers $x_{0}$ and
$x_{1}\in\left[0,1\right]$, their sum $x_{0}+x_{1}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}\right)\subseteq\iota\left(x_{0}\right)+\iota\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
Then, given a three-dimensional Hilbert space, when $\iota$ satisfies
Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add}),
$\iota\circ\mu$ still must be a QIVPM~\citep{HOSTunpublished}.
More concretely, consider two states $\rho_{0}=\frac{1}{3}\proj{0}+\frac{1}{3}\proj{1}+\frac{1}{3}\proj{2}$
and $\rho_{1}=\frac{7}{24}\proj{0}+\frac{1}{3}\proj{1}+\frac{3}{8}\proj{2}$,
and a set of intervals
\begin{equation}
\begin{aligned}\mathscr{I}_{2}={} & \left\{ \imposs,\left[0,\tfrac{1}{4}\right],\left[0,{\textstyle \tfrac{1}{2}}\right],\left[\tfrac{1}{4},\tfrac{3}{4}\right],\right.\\
 & \left.\left[\tfrac{1}{2},1\right],\left[\tfrac{1}{4},1\right],\necess\right\} \,.
\end{aligned}
\end{equation}
Since the interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$
defined in Table~\ref{tab:I_2} satisfies Eqs.~(\ref{eq:iota-constraints})
and Eq.~(\ref{eq:iota-add})~\citep{HOSTunpublished}, $\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$
and $\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$ are both QIVPMs.
\begin{table}
\caption{\label{tab:I_2}A interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$.}
\centering{}%
\begin{tabular}{cc}
\toprule 
\addlinespace
$x$ & $\iota_{2}\left(x\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$x=0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0<x\le\tfrac{1}{8}$ & $\left[0,\tfrac{1}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{1}{8}<x<\tfrac{3}{8}$ & $\left[0,{\textstyle \tfrac{1}{2}}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{3}{8}\le x\le\tfrac{5}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{5}{8}<x<\tfrac{7}{8}$ & $\left[\tfrac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{7}{8}\le x<1$ & $\left[\tfrac{1}{4},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$x=1$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Given any one dimensional projector $\proj{\psi}$, we have $\mu_{\rho_{0}}^{\mathrm{B}}\left(\proj{\psi}\right)=\frac{1}{3}$
so that $\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(\proj{\psi}\right)\right)=\left[0,{\textstyle \tfrac{1}{2}}\right]$.
On the other hand, 
\begin{eqnarray*}
\mu_{\rho_{1}}^{\mathrm{B}}\left(\proj{\psi}\right) & = & \tfrac{7}{24}\left|\ip{\psi}{0}\right|^{2}+\tfrac{1}{3}\left|\ip{\psi}{1}\right|^{2}+\tfrac{3}{8}\left|\ip{\psi}{2}\right|^{2}
\end{eqnarray*}
which is bounded between $\frac{7}{24}$ and $\frac{3}{8}$. Therefore,
$\mu_{\rho_{1}}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$.
According to the complement condition Eq.~(\ref{eq:complement}),
we also have $\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)\in\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)\right)$
for any two dimensional projector~$P$. Hence, $\rho_{1}\in\coreBorn\left(\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}},\events\right)$,
and Thm.~\ref{thm:Finite-precision-Gleason} asserts $\left\Vert \rho_{0}-\rho_{1}\right\Vert $
less than or equal to the maximum length of intervals $\frac{1}{2}$.
This is true because the exact value of $\left\Vert \rho_{0}-\rho_{1}\right\Vert $
is $\frac{1}{24}$.

>>>>

As we have discussed, Statement~\ref{enu:iota-subject-to} implies
\ref{enu:iota-mu-QIVPM}. The rest of the proof is to show that if
$\iota$ is not a THOS rule, then $\iota\circ\mu$ might not be a
QIVPM. Suppose there are three particular numbers~$x_{0}$, $x_{1}$,
and $x_{2}\in\left[0,1\right]$ such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$,
but they don't satisfy Eq.~(\ref{eq:iota-Inclusion}). Consider the
state 
\begin{equation}
\begin{aligned}\rho={} & x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}\\
 & +\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}\,.
\end{aligned}
\end{equation}
whose inducing function $\iota\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}$
fails to satisfy Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}=\proj{0}+\proj{2}$ and $P_{1}=\proj{1}+\proj{2}$, and
cannot be a QIVPM. Therefore, $\iota$ is a THOS rule if and only
if $\iota\circ\mu$ is always a QIVPM. Similarly, Statement~\ref{enu:iota-subject-to}
and \ref{enu:iota-mu-CIVPM} are equivalent as well~\citep{HOSTunpublished}.

In contrast with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM},
there is always a state consistent with any composited QIVPM $\iota\circ\mu_{\rho}^{\mathrm{B}}$.
Moreover, the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower. Eventually, the collection of consisting states collapses
to the singleton set~$\left\{ \rho\right\} $ as the intervals becomes
the infinitely precise intervals~$\mathscr{I}_{\infty}$. This fact
is a weak version of Gleason's theorem because we only consider composited
QIVPMs $\iota\circ\mu_{\rho}^{\mathrm{B}}$ instead of general QIVPMs~$\bar{\mu}$.
To precisely state this result, we need to introduce the norm of pure
states and operators. The norm of a pure state~$\ket{\psi}$ is defined
as usual $\left\Vert \psi\right\Vert =\sqrt{\ip{\psi}{\psi}}$. Similarly,
given any Hermitian operator~$A$, we consider the usual operator
norm
\begin{equation}
\left\Vert A\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert A\ket{\psi}\right\Vert \,,
\end{equation}
which is also known as $2$-norm or spectral norm~\citep{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.
Then, a weak version of Gleason's theorem can be stated as follow.

@@

Given a state~$\rho'$ consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$,
we have $\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota\left(\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$
for any one-dimensional projector $\proj{\psi}$. Since the maximum
length of the intervals in $\mathscr{I}$ is $\delta$, it is also
the upper bound of the difference 
\[
\left|\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right|=\left|\melem{\psi}{\rho-\rho'}{\psi}\right|\le\delta\,.
\]
Since $\rho-\rho'$ is Hermitian, $\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$
is the maximum absolute value of the eigenvalues of $\rho-\rho'$~\citep{544199},
and equal to $\left\Vert \rho-\rho'\right\Vert $~\citep{GolubVanLoan1996,Foucart2012}.
Hence, $\left\Vert \rho-\rho'\right\Vert \le\delta$.

@@

Even if $\iota$ has to be a THOS rule to guarantee $\iota\circ\mu$
a QIVPM, the composited function $\iota\circ\mu$ may still be a QIVPM
when $\iota$ is not a THOS rule. In this case, the composited function
$\iota\circ\mu$ could map to only finite number of intervals. For
example, we can weaken the convex condition, Eq.~(\ref{eq:iota-Inclusion}),
to the following additive condition: for any numbers $x_{0}$ and
$x_{1}\in\left[0,1\right]$, their sum $x_{0}+x_{1}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}\right)\subseteq\iota\left(x_{0}\right)+\iota\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
Then, given a three-dimensional Hilbert space, when $\iota$ satisfies
Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add}),
$\iota\circ\mu$ still must be a QIVPM~\citep{HOSTunpublished}.
More concretely, consider two states $\rho_{0}=\frac{1}{3}\proj{0}+\frac{1}{3}\proj{1}+\frac{1}{3}\proj{2}$
and $\rho_{1}=\frac{7}{24}\proj{0}+\frac{1}{3}\proj{1}+\frac{3}{8}\proj{2}$,
and a set of intervals
\begin{equation}
\begin{aligned}\mathscr{I}_{2}={} & \left\{ \imposs,\left[0,\tfrac{1}{4}\right],\left[0,{\textstyle \tfrac{1}{2}}\right],\left[\tfrac{1}{4},\tfrac{3}{4}\right],\right.\\
 & \left.\left[\tfrac{1}{2},1\right],\left[\tfrac{1}{4},1\right],\necess\right\} \,.
\end{aligned}
\end{equation}
Since the interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$
defined in Table~\ref{tab:I_2} satisfies Eqs.~(\ref{eq:iota-constraints})
and Eq.~(\ref{eq:iota-add})~\citep{HOSTunpublished}, $\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$
and $\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$ are both QIVPMs.
\begin{table}
\caption{\label{tab:I_2}A interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$.}

\hfill{}%
\begin{minipage}[t]{2.5cm}%
\begin{center}
\begin{tabular}{cc}
\toprule 
\addlinespace
$x$ & $\iota_{2}\left(x\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$x=0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0<x\le\tfrac{1}{8}$ & $\left[0,\tfrac{1}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{1}{8}<x<\tfrac{3}{8}$ & $\left[0,{\textstyle \tfrac{1}{2}}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{3}{8}\le x\le\tfrac{5}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\par\end{center}%
\end{minipage}\hfill{}%
\begin{minipage}[t]{2.5cm}%
\begin{center}
\begin{tabular}{cc}
\toprule 
\addlinespace
$x$ & $\iota_{2}\left(x\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\tfrac{5}{8}<x<\tfrac{7}{8}$ & $\left[\tfrac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{7}{8}\le x<1$ & $\left[\tfrac{1}{4},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$x=1$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\par\end{center}%
\end{minipage}\hfill{}
\end{table}
Given any one-dimensional projector $\proj{\psi}$, we have $\mu_{\rho_{0}}^{\mathrm{B}}\left(\proj{\psi}\right)=\frac{1}{3}$
so that $\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(\proj{\psi}\right)\right)=\left[0,{\textstyle \tfrac{1}{2}}\right]$.
On the other hand, 
\begin{eqnarray*}
\mu_{\rho_{1}}^{\mathrm{B}}\left(\proj{\psi}\right) & = & \tfrac{7}{24}\left|\ip{\psi}{0}\right|^{2}+\tfrac{1}{3}\left|\ip{\psi}{1}\right|^{2}+\tfrac{3}{8}\left|\ip{\psi}{2}\right|^{2}
\end{eqnarray*}
which is bounded between $\frac{7}{24}$ and $\frac{3}{8}$. Therefore,
$\mu_{\rho_{1}}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$.
According to the complement condition, Eq.~(\ref{eq:complement}),
we also have $\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)\in\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)\right)$
for any two-dimensional projector~$P$. Hence, $\rho_{1}\in\coreBorn\left(\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}},\events\right)$,
and Thm.~\ref{thm:Finite-precision-Gleason} asserts $\left\Vert \rho_{0}-\rho_{1}\right\Vert $
less than or equal to the maximum length of intervals $\frac{1}{2}$.
This is true because the exact value of $\left\Vert \rho_{0}-\rho_{1}\right\Vert $
is $\frac{1}{24}$.
